---
title: "LFS_Algorithms"
author: "Azad"
date: "2023-11-05"
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
---

# Define the file path and Read CSV
```{r}
full_dataset_path <- "C:/820WF/LFS_2017_2023/lfs_17_23.csv"
LFS_data <- read.csv(full_dataset_path)
```

# Select a smaller subset of data and Combine Categories
```{r}
# Select only the relevant variables
LFS_Selected <- LFS_data[LFS_data$SURVYEAR >= 2023, c("LFSSTAT", "PROV", "AGE_12", "SEX", "MARSTAT", "EDUC", "IMMIG")]

```


#Dependent and Independent Variables
```{r}
##Dependent Variable: LFSSTAT

# LFSSTAT (Labour force status) contains the following:
#1	Employed, at work
#2	Employed, absent from work
#3	Unemployed
#4	Not in labour force

#For the purpose of data analysis the dependent variable is combined into two levels
# 1 = 1 = Active Employment
# 2 = 2,3,4 = Inactive employment

## Independent variables: Six variables taken as independent variables:

# prov =	Province
# age_12 =	Five-year age group of respondent
# sex	= Sex of respondent
# marstat =	Marital status of respondent
# educ =	Highest educational attainment
# immig =	Immigrant status


#Algorithms

# The algorithms that are used are Decision Tree, Random Forest and Naive Bayes
# Cross-validation is used to assess the performance of the models
# Accuracy, Precision and Recalls are used to see how the models performed

```


# Combine the four LFSSTAT categories into two levels
```{r}
LFS_Selected$LFSSTAT <- ifelse(LFS_Selected$LFSSTAT == 1, 1, 2)
```


# Load the Libraries
```{r}
library(randomForest)
library(caret)
library(rpart)
library(e1071)
```

# Algorithm 1: Decision Tree
```{r}
# Prepare the Data
LFS_Selected$LFSSTAT <- as.factor(LFS_Selected$LFSSTAT)

# cross-validation settings
control_ndt <- trainControl(method = "cv", number = 5)  # 5-fold cross-validation

# Create a smaller training subset
set.seed(123)

sample_size <- floor(0.1 * nrow(LFS_Selected))  # Use 10% of the data
train_index_ndt <- sample(1:nrow(LFS_Selected), sample_size)
smaller_training_data_ndt <- LFS_Selected[train_index_ndt, ]
testing_data_ndt <- LFS_Selected[-train_index_ndt, ]

# Train a Decision Tree Model with cross-validation
tree_model_ndt <- train(LFSSTAT ~ ., data = smaller_training_data_ndt, method = "rpart", trControl = control_ndt)

# Make Predictions with class labels
tree_predictions_class <- predict(tree_model_ndt, testing_data_ndt)

# confusion matrix
confusion_tree <- confusionMatrix(tree_predictions_class, testing_data_ndt$LFSSTAT)

# Accuracy, precision and recall
accuracy_tree <- confusion_tree$overall['Accuracy']
precision_tree <- confusion_tree$byClass['Pos Pred Value']
recall_tree <- confusion_tree$byClass['Sensitivity']

# Print the confusion matrix and results
print(confusion_tree)
cat("Accuracy:", accuracy_tree, "\n")
cat("Precision:", precision_tree, "\n")
cat("Recall:", recall_tree, "\n")
```
#Visualize Decision Tree Confusion Matrix
```{r}
# Convert the confusion matrix to a data frame for plotting
conf_matrix_tree_df <- as.data.frame(as.table(confusion_tree))

# Calculate percentages for a more informative plot
conf_matrix_tree_df$Percentage <- conf_matrix_tree_df$Freq / sum(conf_matrix_tree_df$Freq) * 100

# Plot the confusion matrix using ggplot2
conf_plot_dt <- ggplot(conf_matrix_tree_df, aes(x = Reference, y = Prediction, fill = Percentage)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.1f%%", Percentage)), vjust = 1) +
  scale_fill_gradient(low = "white", high = "skyblue") +  # Adjust the color scale
  theme_minimal() +
  labs(title = "Confusion Matrix - Decision Tree",
       x = "Actual",
       y = "Predicted") +
  theme(axis.text = element_text(size = 12),       # Adjust text size
        axis.title = element_text(size = 14),      # Adjust axis title size
        plot.title = element_text(size = 16))      # Adjust plot title size

# Display the plot
print(conf_plot_dt)


```



# Algorithm 2: Random Forest
```{r}
# Split the dataset into a training set and a test set
set.seed(123)
trainIndex <- createDataPartition(LFS_Selected$LFSSTAT, p = 0.7, list = FALSE)

# the data set is large so the dataset is shrinked into a smaller one
smaller_training_data <- LFS_Selected[sample(nrow(LFS_Selected), nrow(LFS_Selected) / 10), ]

# the testing data
testing_data <- LFS_Selected[-trainIndex, ]

# cross-validation
control_rf <- trainControl(method = "cv", number = 10)  # 10-fold cross-validation

# Train the Random Forest model using cross-validation
model_rf <- train(factor(LFSSTAT) ~ ., data = smaller_training_data, method = "rf", trControl = control_rf, ntree = 25)

# Make predictions using the model
predictions_rf <- predict(model_rf, testing_data)

# Ensure both are factors with the same levels
predictions_rf <- as.factor(predictions_rf)
testing_data$LFSSTAT <- as.factor(testing_data$LFSSTAT)

# Confusion matrix
confusion_rf <- confusionMatrix(predictions_rf, testing_data$LFSSTAT)

# Calculate accuracy, precision, and recall
accuracy_rf <- confusion_rf$overall['Accuracy']
precision_rf <- confusion_rf$byClass['Pos Pred Value']
recall_rf <- confusion_rf$byClass['Sensitivity']

# Print the confusion matrix and results
print(confusion_rf)
cat("Accuracy:", accuracy_rf, "\n")
cat("Precision:", precision_rf, "\n")
cat("Recall:", recall_rf, "\n")
```
#Visualize Random Forest Confusion Matrix
```{r}
# Convert the confusion matrix to a data frame for plotting
conf_matrix_rf_df <- as.data.frame(as.table(confusion_rf))

# Calculate percentages for a more informative plot
conf_matrix_rf_df$Percentage <- conf_matrix_rf_df$Freq / sum(conf_matrix_rf_df$Freq) * 100

# Plot the confusion matrix using ggplot2
conf_plot_rf <- ggplot(conf_matrix_rf_df, aes(x = Reference, y = Prediction, fill = Percentage)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.1f%%", Percentage)), vjust = 1) +
  scale_fill_gradient(low = "white", high = "skyblue") +  # Adjust the color scale
  theme_minimal() +
  labs(title = "Confusion Matrix - Random Forest",
       x = "Actual",
       y = "Predicted") +
  theme(axis.text = element_text(size = 12),       # Adjust text size
        axis.title = element_text(size = 14),      # Adjust axis title size
        plot.title = element_text(size = 16))      # Adjust plot title size

# Display the plot
print(conf_plot_rf)


```


## Algorithm 3: Naive Bayes
```{r}
LFS_Selected$LFSSTAT <- as.factor(LFS_Selected$LFSSTAT)

# cross-validation
control_nb <- trainControl(method = "cv", number = 5)  # 5-fold cross-validation

# Create a smaller training subset
set.seed(123)
sample_size <- floor(0.1 * nrow(LFS_Selected))  # Use 10% of the data
train_index_nb <- sample(1:nrow(LFS_Selected), sample_size)
smaller_training_data_nb <- LFS_Selected[train_index_nb, ]
testing_data_nb <- LFS_Selected[-train_index_nb, ]

# Train a Naive Bayes Model with cross-validation
nb_model <- train(LFSSTAT ~ ., data = smaller_training_data_nb, method = "naive_bayes", trControl = control_nb)

# Make Predictions
nb_predictions <- predict(nb_model, testing_data_nb)

# confusion matrix
confusion_nb <- confusionMatrix(nb_predictions, testing_data_nb$LFSSTAT)

# Calculate accuracy, precision, and recall
accuracy_nb <- confusion_nb$overall['Accuracy']
precision_nb <- confusion_nb$byClass['Pos Pred Value']
recall_nb <- confusion_nb$byClass['Sensitivity']

# Print the confusion matrix and results
print(confusion_nb)

cat("Accuracy:", accuracy_nb, "\n")
cat("Precision:", precision_nb, "\n")
cat("Recall:", recall_nb, "\n")
```
#Visualize Naive Bayes Confusion Matrix
```{r}
# Convert the confusion matrix to a data frame for plotting
conf_matrix_nb_df <- as.data.frame(as.table(confusion_nb))

# Calculate percentages for a more informative plot
conf_matrix_nb_df$Percentage <- conf_matrix_nb_df$Freq / sum(conf_matrix_nb_df$Freq) * 100

# Plot the confusion matrix using ggplot2
conf_plot_nb <- ggplot(conf_matrix_nb_df, aes(x = Reference, y = Prediction, fill = Percentage)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.1f%%", Percentage)), vjust = 1) +
  scale_fill_gradient(low = "white", high = "skyblue") +  # Adjust the color scale
  theme_minimal() +
  labs(title = "Confusion Matrix - Naive Bayes",
       x = "Actual",
       y = "Predicted") +
  theme(axis.text = element_text(size = 12),       # Adjust text size
        axis.title = element_text(size = 14),      # Adjust axis title size
        plot.title = element_text(size = 16))      # Adjust plot title size

# Display the plot
print(conf_plot_nb)


```


# LFS Oct: Read October 2023 Dataset
```{r}
Oct_dataset_path <- "C:/820WF/LFS_2017_2023/lfs_1023.csv"
Oct_data <- read.csv(Oct_dataset_path)
```

# LFS Oct: Select a smaller subset of data and Combine Categories
```{r}
# Select only the relevant variables
Oct_Selected <- Oct_data[c("SURVYEAR","LFSSTAT", "PROV", "AGE_12", "SEX", "MARSTAT", "EDUC", "IMMIG")]

# Combine the four LFSSTAT categories into two levels
Oct_Selected$LFSSTAT <- ifelse(Oct_Selected$LFSSTAT == 1, 1, 2)
```

# LFS Oct: Predictions on Oct Data
```{r}
# Make predictions using the model
predictions_rf_Oct <- predict(model_rf, Oct_Selected)

# Ensure both are factors with the same levels
predictions_rf_Oct <- as.factor(predictions_rf_Oct)
Oct_Selected$LFSSTAT <- as.factor(Oct_Selected$LFSSTAT)

# Confusion matrix
confusion_rf_Oct <- confusionMatrix(predictions_rf_Oct, Oct_Selected$LFSSTAT)

# Calculate accuracy, precision, and recall
accuracy_rf_Oct <- confusion_rf_Oct$overall['Accuracy']
precision_rf_Oct <- confusion_rf_Oct$byClass['Pos Pred Value']
recall_rf_Oct <- confusion_rf_Oct$byClass['Sensitivity']

# Print the confusion matrix and results
print(confusion_rf_Oct)
cat("Accuracy:", accuracy_rf_Oct, "\n")
cat("Precision:", precision_rf_Oct, "\n")
cat("Recall:", recall_rf_Oct, "\n")

```

#Visualize Random Forest Confusion Matrix - Oct
```{r}
# Convert the confusion matrix to a data frame for plotting
conf_matrix_rf_Oct_df <- as.data.frame(as.table(confusion_rf_Oct))

# Calculate percentages for a more informative plot
conf_matrix_rf_Oct_df$Percentage <- conf_matrix_rf_Oct_df$Freq / sum(conf_matrix_rf_Oct_df$Freq) * 100

# Plot the confusion matrix using ggplot2
conf_plot_rf_Oct <- ggplot(conf_matrix_rf_Oct_df, aes(x = Reference, y = Prediction, fill = Percentage)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.1f%%", Percentage)), vjust = 1) +
  scale_fill_gradient(low = "white", high = "skyblue") +  # Adjust the color scale
  theme_minimal() +
  labs(title = "Confusion Matrix - Random Forest",
       x = "Actual",
       y = "Predicted") +
  theme(axis.text = element_text(size = 12),       # Adjust text size
        axis.title = element_text(size = 14),      # Adjust axis title size
        plot.title = element_text(size = 16))      # Adjust plot title size

# Display the plot
print(conf_plot_rf_Oct)


```
